---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```


### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv
```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```



### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024`-`Fall 2021`)/ `Fall 2021` * 100)
```


### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?


In the data, I found that the university increased most fees from 2021 to 2024. One of the fees that increased the most was student sustainability, with a 150 percent increase from 2021 to 2024. For a story, I would find out if students were aware of this change and what is actually funded through these fees. Is this money actually being given back to students through improved experiences at school or not? Were students ever actually made aware of this change or did it just show up on the bill with no notice?


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
