---
title: "MidtermNotes"
output: html_document
---

### Homework 1
# load tidyverse
```{r}
library(tidyverse)
```
# make a dataframe
```{r}
students <- read_csv("link")
```

### Homework 2
# calculating average
```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)

pg_crime_averages <- pg_crime_averages |> mutate(average_incidents = (Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)/ 12)
```
# calculating percent change
```{r}
summer_crime_averages <- pg_crime_averages |> mutate(
  june_change = (Jun - average_incidents)/average_incidents *100,
july_change = (Jul - average_incidents)/average_incidents *100,
august_change = (Aug - average_incidents)/average_incidents *100 )
```
### Homework 3
# mean, median, minimum, maximum
```{r}
wh_salaries |> summarize(
    mean_value = mean(salary),
    median_value = median(salary),
     min_value = min(salary),
    max_value = max(salary)
  )
```
### Homework 4
# load the janitor library
```{r}
library(janitor)
```
#calculating rates
```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 *1000, violent_rate_2020 = violent2020/pop2020 * 1000, property_rate_2020 = property2020/pop2020 *1000)
```
### Homework 5
#percentile
```{r}
md_cities_percentiles <- md_cities_change |> mutate(percentile = ntile(violent_rate_2020, 100))
```
#IQR
```{r}
md_cities_percentiles |> summarize(interquartile = IQR(violent_rate_2020)) 
```
### Homework 6
#mean and standard deviation
```{r}
spotify_songs |> summarize(mean = mean(danceability), sd = sd(danceability) )
```
# histogram
```{r}
spotify_songs |> 
   ggplot() +
  geom_histogram(aes(x=danceability), binwidth = .1) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed" , size=1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed" , size=1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed" , size=1)
```
### Homework 7
#sampling
```{r}
sample5 <- candy |> sample_n(5)

sample10 <- candy |> sample_n(10)

sample5 |> summarize(mean = mean(red), sd = sd(red))

sample10 |> summarize(mean = mean(red), sd = sd(red))
```
### Homework 8
# correlation coefficient
```{r}
gdp_data |> summarize(correlation = cor(GDP, DJIA, method = "pearson"))
```
# scatterplot
```{r}
gdp_data |> 
  ggplot() +
  geom_point(aes(x=DJIA, y=GDP)) +
  geom_smooth(aes(x=DJIA, y=GDP), method = "lm") +
  ggtitle("GDP vs DJIA Per Quarter")
```

you can also use labs(title = )
