
### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")

```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarize(mean = mean(danceability), sd = sd(danceability) )
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> 
   ggplot() +
  geom_histogram(aes(x=danceability), binwidth = .1)

```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> 
   ggplot() +
  geom_histogram(aes(x=danceability), binwidth = .1) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed" , size=1)

```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs |> 
   ggplot() +
  geom_histogram(aes(x=danceability), binwidth = .1) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed" , size=1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed" , size=1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed" , size=1)

```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The distribution is somewhat normal as you can see a slight bell curve, but I would definitely describe it as skewed to the right. The data that is most frequently occurring ranges from around .5 to .875 in terms of danceability scores. The highest occurring danceability score is around .69, which is very close to to the mean. 

Since the mean and median are so similar, the story can center around the fact that songs on average have a danceability score of about .67, which is interesting because such a large part of listening to music is dancing. For this story I would talk to people about how often they dance to music and why, and then compare this to the mean. I think this average danceability score is a little low, but I'd be interested to see through interviews what people think.



### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

