---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```



### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
candy <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```



### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample5 <- candy |> sample_n(5)

sample10 <- candy |> sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 |> summarize(mean = mean(red), sd = sd(red))

sample10 |> summarize(mean = mean(red), sd = sd(red))
```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
candy |> summarize(mean = mean(red), sd = sd(red))
```



### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The means are all pretty similar, with the mean of the population being 72 and the means of the samples being around 75. The standard deviations differ a little bit with the standard deviation of the population and sample10 being 16, but the sample5 is 5.95. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

I found that the sample10 mean and sd were more indicative of the mean and sd of the population. This means random sampling can lead to inaccurate data in terms of the mean and standard deviation of a population, especially if the sample is smaller. Using a larger random sample will ensure that the mean and standard deviation of the sample more accurately reflect the population as a whole. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
