---
title: "final_project"
output: html_document
---

Links to data:

men-
https://en.wikipedia.org/wiki/United_States_men%27s_national_soccer_team_results

women-
https://en.wikipedia.org/wiki/United_States_women%27s_national_soccer_team_results

```{r}
library(tidyverse)
```
```{r}
usmnt_games <- read_csv("USMNT GAME SCORES - Sheet1.csv")
```
```{r}
uswnt_games <- read_csv("USWNT GAMES - Sheet1.csv")
```

Description of data: These dataframes include the match number, opponent, date, result and event type for every USMNT match from 1916 to 2025 and every USWNT match from 1985 to 2025.For my analysis, the most important column will most likely be the score result, as my project will focus on comparing the two teams goal differentials. If I choose to do further analysis on the strength of the opponent or the seriousness of the games, the opponent and event columns will be important too. I will need to add columns that specify goals scored and goals allowed as well as if the match was a knockout game or if the opponent was ranked highly at the time. 
